Penalized regression for left-truncated and right-censored survival data

### 2021-04-30 manuscript.tex
footnote change
minor rewordings

\begin{figure}[h]
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/followup.pdf} 
\caption{Left-truncated and right-censored patient follow-up in a hypothetical study cohort, ordered chronologically by event time. Patients who receive a diagnosis (closed circle) become eligible to enter the cohort after reaching a milestone (black triangle), for example a genomic test. Patients are followed until death (open circle) or censoring (cross). However, patients who die before reaching the milestone are left-truncated (in red), and only those who have survived until eligibility (in black) are observed. Left truncation time, or the time between diagnosis and cohort entry, is shown with a dashed line. Left truncation time is also referred to as ``entry time.''}
\label{fig:followup}
\end{figure}

@@

\subsection{Real-world use case}
We consider the Clinico-Genomic Database offered jointly by Flatiron Health and Foundation Medicine (FH-FMI CGDB) \cite{singal2017cgdb}. The CGDB is a US-based de-identified oncology database that combines real-world, patient-level clinical data and outcomes with patient-level genomic data to provide a comprehensive patient and tumor profile. The de-identified data originated from approximately 280 US cancer clinics (~800 sites of care). Patients in the CGDB have received at least one Foundation Medicine comprehensive genomic profiling test as well as have had their electronic health records captured by Flatiron Health. The retrospective, longitudinal EHR-derived database comprises patient-level structured and unstructured data, curated via technology-enabled abstraction, and were linked to Foundation Medicine genomic data by de-identified, deterministic matching. Genomic alterations were identified via comprehensive genomic profiling (CGP) of >300 cancer-related genes on FMI's next-generation sequencing (NGS) based FoundationOne panel \cite{birnbaum2020modelassisted, ma2020fmi}. To date, over 400,000 samples have been sequenced from patients across the US. The data are de-identified and subject to obligations to prevent re-identification and protect patient confidentiality. Altogether, the CGDB represents thousands of potential predictors per patient, ranging from demographic, laboratory, and treatment information to the specific mutations detected on each biomarker included in a Foundation Medicine baitset. Institutional Review Board approval of the study protocol was obtained prior to study conduct, and included a waiver of informed consent.

@@

\begin{table}[!ht]
\begin{center}
\begin{threeparttable}
\caption{Comparison of model discrimination in the CGDB} \label{tbl:model-discrimination}
\begin{tabularx}{\textwidth}{@{\extracolsep{\fill}}llrr}
\hline
\multicolumn{2}{l}{} & \multicolumn{2}{c}{C-index}  \\
\cmidrule(r){3-4}  
\multicolumn{1}{l}{Model} & \multicolumn{1}{l}{Covariate size} & \multicolumn{1}{l}{No adjustment} & \multicolumn{1}{l}{Adjustment}  \\
\hline
\ExpandableInput{tables/performance.txt}
\hline
\end{tabularx}
\scriptsize Note: ``Adjusted'' models incorporate left truncation times.
\end{threeparttable}
\end{center}
\end{table}

@@

The adjusted models tend to be better calibrated for patients with the lowest survival probabilities but poorly calibrated for patients with higher survival probabilities. One possible reason for this result is that the adjusted models pushed many of the hazard ratios to $1$ and did not adequately separate low and high risk patients. Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso.

@@

In our real-world use case, an analyst could have mistakenly concluded that the high C-index in the non-adjusted model was indicative of good model performance when it was in fact driven primarily by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor ``year of diagnosis'' moved from being the largest in absolute value to around 1 after adjusting for left truncation. The strong positive relationship between ``year of diagnosis'' on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias. This relationship reversed completely after adjusting for left truncation, and illustrates that biases to coefficient estimates are driven by the association between predictors and left truncation time. In the case of ``year of diagnosis'', this association was negative and biased the hazard ratio upwards. While many of the other hazard ratios in the large $p$ model were pushed towards 1 after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor. In cases where predictors and left truncation time are positively associated, hazard ratios are biased downwards because the predictors are associated with more immortal time. Because the high C-index misleadingly suggested good model performance, we recommend that models should be evaluated using both discrimination and calibration based metrics to facilitate interpretation.

>>>>

\begin{figure}[h]
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/followup.pdf} 
\caption{Left-truncated and right-censored patient follow-up in a hypothetical study cohort, ordered chronologically by event time. Patients who receive a diagnosis (closed circle) become eligible to enter the cohort after reaching a milestone (black triangle), for example a genomic test. Patients are followed until death (open circle) or censoring (cross). However, patients who die or are censored before reaching the milestone are left-truncated (in red), and only those who have survived until eligibility (in black) are observed. Left truncation time, or the time between diagnosis and cohort entry, is shown with a dashed line. Left truncation time is also referred to as ``entry time.''}
\label{fig:followup}
\end{figure}

@@

\subsection{Real-world use case}
We consider the Clinico-Genomic Database offered jointly by Flatiron Health and Foundation Medicine (FH-FMI CGDB) \cite{singal2017cgdb}. The CGDB is a US-based de-identified oncology database that combines real-world, patient-level clinical data and outcomes with patient-level genomic data to provide a comprehensive patient and tumor profile. The de-identified data originated from approximately 280 US cancer clinics (roughly 800 sites of care). Patients in the CGDB have received at least one Foundation Medicine comprehensive genomic profiling test as well as have had their electronic health records captured by Flatiron Health. The retrospective, longitudinal EHR-derived database comprises patient-level structured and unstructured data, curated via technology-enabled abstraction, and were linked to Foundation Medicine genomic data by de-identified, deterministic matching. Genomic alterations were identified via comprehensive genomic profiling (CGP) of >300 cancer-related genes on FMI's next-generation sequencing (NGS) based FoundationOne panel \cite{birnbaum2020modelassisted, ma2020fmi}. To date, over 400,000 samples have been sequenced from patients across the US. The data are de-identified and subject to obligations to prevent re-identification and protect patient confidentiality. Altogether, the CGDB represents thousands of potential predictors per patient, ranging from demographic, laboratory, and treatment information to the specific mutations detected on each biomarker included in a Foundation Medicine baitset. Institutional Review Board approval of the study protocol was obtained prior to study conduct, and included a waiver of informed consent.

@@

\begin{table}[!ht]
\begin{center}
\begin{threeparttable}
\caption{Comparison of model discrimination in the CGDB} \label{tbl:model-discrimination}
\begin{tabularx}{\textwidth}{@{\extracolsep{\fill}}llrr}
\hline
\multicolumn{2}{l}{} & \multicolumn{2}{c}{C-index}  \\
\cmidrule(r){3-4}  
\multicolumn{1}{l}{Model} & \multicolumn{1}{l}{Covariate size} & \multicolumn{1}{l}{No adjustment} & \multicolumn{1}{l}{Adjustment}  \\
\hline
\ExpandableInput{tables/performance.txt}
\hline
\end{tabularx}
\scriptsize Note: ``Adjustment'' indicates models with a left truncation adjustment.
\end{threeparttable}
\end{center}
\end{table}

@@

The adjusted models tend to be better calibrated for patients with the lowest survival probabilities but poorly calibrated for patients with higher survival probabilities. One possible reason for this result is that the adjusted models pushed many of the hazard ratios to $1$ and did not adequately separate low and high risk patients. Probability rescaling such as Platt scaling or isotonic regression may be effective on the adjusted probabilities here, which resemble the characteristic sigmoid-shaped calibration curves produced by different learning algorithms, including the lasso; however the implementation is less straightforward in a survival context \cite{goldstein2021xcal}.

@@

In our real-world use case, an analyst could have mistakenly concluded that the C-index in the non-adjusted model was indicative of good model performance when it was in fact driven in part by a high correlation between predictors and left truncation. For instance, in our large $p$ model the hazard ratio for the predictor ``year of diagnosis'' moved from being the largest in absolute value to around 1 after adjusting for left truncation. The strong positive relationship between ``year of diagnosis'' on the hazard in the non-adjusted model was artificial because patients diagnosed earlier appeared to survive longer due to immortal time bias. This relationship reversed completely after adjusting for left truncation, and illustrates that biases to coefficient estimates are driven by the association between predictors and left truncation time. In the case of ``year of diagnosis'', this association was negative and biased the hazard ratio upwards. While many of the other hazard ratios in the large $p$ model were pushed towards 1 after left truncation adjustment, the direction of this bias depends on the direction of the association with left truncation time and varies by predictor. In cases where predictors and left truncation time are positively associated, hazard ratios are biased downwards because the predictors are associated with more immortal time. Because the high C-index misleadingly suggested good model performance, we recommend that models should be evaluated using both discrimination and calibration based metrics to facilitate interpretation.
